A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-08 01-intro.tex
Pequenas revisões.

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language for embedded soft real-time systems that aims to offer a
concurrent, safe, and expressive alternative to C with the characteristics that
follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming let developers write code in direct/sequential
style, recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
Synchronous languages offer a simple run-to-completion model of execution that
enable deterministic execution and make formal reasoning tractable.
For this reason, it has been successfully adopted in safety-critical real-time
embedded systems.~\cite{rp.twelve}

Previous work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\% for \emph{text} and \emph{data})~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}, and as an
alternative language in an undergraduate-level course on embedded systems for
the past 6 years.

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control.~\cite{ceu.tecs17}
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
    \item Stack-based execution for internal events, which provides a limited
          form of coroutines.
    \item Fine-grained, intra-reaction deterministic execution, which allows
          programs to safely share memory.
    \item Finalization mechanism for abortion of lines of execution, which
          safely release external resources.
    \item First-class synchronized timers.
\end{itemize}

>>>>

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language for embedded soft real-time systems that aims to offer a
concurrent, safe, and expressive alternative to C with the characteristics that
follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming let developers write code in direct/sequential
style, recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
Synchronous languages offer a simple run-to-completion model of execution that
enable deterministic execution and make formal reasoning tractable.
For this reason, it has been successfully adopted in safety-critical real-time
embedded systems~\cite{rp.twelve}.

Previous work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\% for \emph{text} and \emph{data})~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.
%, and as an
%alternative language in an undergraduate-level course on embedded systems for
%the past 6 years.

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control.~\cite{ceu.tecs17}
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regards to external resources.
\end{itemize}

### 2018-02-08 02-ceu.tex
Pequenas revisões.

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
% TODO: or every
%
This restriction, which is statically checked by the compiler, ensures that
every reaction runs in bounded time, eventually terminating with all trails
blocked in \code{await} statements.
%
\CEU has an additional restriction, which it shares with Esterel and
synchronous languages in general~\cite{esterel.preemption}: computations that
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

@@

At first sight, event iteration such as ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (\code{e} is an internal
event).
Event iterators have syntactic restrictions and cannot contain \code{await} or
\code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as safe blocking point in the program.

\begin{comment}
%
Third, \CEU subroutines can be safely used by the other primitives of the
language, such as parallel compositions and the \code{await} statement,
without breaking the programming model.
%
In particular, after calling a subroutine these primitives wait while
keeping context information, such as locals and the program counter, which
makes the calls behave similarly to those of
coroutines~\cite{lua.coroutines}.
%
Finally, in previous work, we built other advanced control mechanisms on top
of internal events, such as resumable exceptions and reactive
variables~\cite{ceu.rem13}.
%
\end{comment}

>>>>

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
%
This restriction, which is statically checked by the compiler, ensures that
every reaction runs in bounded time, eventually terminating with all trails
blocked in \code{await} statements.
%
\CEU has an additional restriction, which it shares with Esterel and
synchronous languages in general~\cite{esterel.preemption}: computations that
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

@@

At first sight, event iteration such as ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (\code{e} is an internal
event).
%
Event iterators have syntactic restrictions and cannot contain \code{await} or
\code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as safe blocking point in the program.
%
For this reason, the restriction that execution paths within loops must
contains at least one external \code{await} is extended to contain an
\code{every} statement alternatively.

### 2018-02-08 03-semantics.tex
Pequenas revisões.

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from the input event, how the program behaves and 
becomes idle again to proceed to the subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathds{N}
    && (current~stack~level)
\\
e, e' &\in \mathds{E \cup \{\epsilon\}}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event, the triple is initialized at
stack level $0$ ($n=0$) and with the emitted event $id$ ($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after a number of transitions, the triple will block
with a (possibly) modified program $p'$, at stack level $0$, with no event
emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also define two transition rule subtypes $\OUT$ (for \emph{outermost}) and
$\NST$ (for \emph{nested}).

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that manipulate the stack level as follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level,
    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and does not apply if the program is not blocked (see $isblocked(n,p)$
in Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes, as discussed in Section~\ref{sec.ceu.evts}.

>>>>

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from the input event, how the program behaves and 
becomes idle again to proceed to the subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathds{N}
    && (current~stack~level)
\\
e, e' &\in \mathds{E \cup \{\epsilon\}}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event $id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event $id$
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program $p'$, at stack level
$0$, with no event emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also distinguish between two transition rule subtypes:
    $\OUT$, for \emph{outermost} transitions; and
    $\NST$, for \emph{nested} transitions.

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level as
follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level,
    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and does not apply if the program is not blocked (see $isblocked(n,p)$
in Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes, as discussed in Section~\ref{sec.ceu.evts}.
